\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}


I want to express my heartfelt gratitude to everyone who contributed, both directly and indirectly, to the completion of this work.

First and foremost, I am profoundly thankful to the professors who have guided me throughout my academic journey. They have played a crucial role in my intellectual and personal development.
I extend my deep appreciation to the INESC TEC team, and particularly to Eduardo Rodrigues, for their invaluable guidance, collaboration, and unwavering support throughout the course of this research.

Additionally, I acknowledge the presence and contributions of all the students who have passed through this faculty. Their shared experiences fostered a stimulating academic and social environment.
A special note of recognition goes to the Academic Traditions, which I wholeheartedly embraced. These traditions significantly enriched my university experience and deepened my sense of belonging to this academic community.

I am immensely grateful to my girlfriend for her steadfast support, patience, and encouragement during the most challenging moments of my academic journey.
To my friends who stood by me over the past five years, providing companionship, motivation, and balance: thank you. I leave with my heart full of our long conversations, unforgettable memories, and the strength drawn from both the joyful and difficult times we shared together.
Finally, I express my deep appreciation to my family, whose unwavering support has always been my foundation. I am especially grateful to my father, whose words have consistently reminded me of the value of resilience: “A vida é dura para quem é mole”

The fight against cancer, which serves as the primary motivation for this research, arises from many years of personal struggle and progress in my health journey. I am profoundly grateful to have reached a point where I can contribute, albeit in a modest way, to alleviating the hardship faced by families impacted by this disease.

This work would not have been achievable without the unwavering support of numerous individuals. Success is never a solitary endeavour, and this dissertation stands as a testament to that truth.

As the Orfeão Universitário do Porto sings: "Quero ficar sempre estudante"\\
With love,\\
Engineering!

\vspace{10mm}
\flushleft{João Malva}

\chapter*{Institutional Acknowledgements}

This work is co-financed by Component 5 - Capitalization and Business Innovation, integrated in the Resilience Dimension of the Recovery and Resilience Plan within the scope of the Recovery and Resilience Mechanism~(MRR) of the European Union~(EU), framed in the Next Generation EU, for the period 2021 - 2026, within project HfPT, with reference 41.

\vspace{10mm}
\flushleft{João Malva}
